---
title: "BRFSS 2022 - Diabetes by race for working-age population in TX"
author: "Kaitlan"
format: html
editor: visual
---

Data downloaded from CDC''s BRFSS 2022 public data files: <https://www.cdc.gov/brfss/annual_data/annual_2022.html/>

**Step 1: Load data. Rename column names and filter for Texas.**

```{r}
# Load the required libraries
library(haven)
library(dplyr)
library(survey)

# Read the XPT file
data <- read_xpt("LLCP2022.XPT")

# Rename columns for convenience
data <- data %>%
  rename(
    STATE = `_STATE`,
    STSTR = `_STSTR`,
    LLCPWT = `_LLCPWT`,
    BMI5CAT = `_BMI5CAT`,
    RACE1 = `_RACE1`,
    AGE65YR = `_AGE65YR`
  )

# Filter for Texas
data_tx <- data %>% filter(STATE == 48) # 48 is the FIPS code for Texas
```

**Step 2: Prep the data by filtering for working-age only and recoding variables.**

```{r}

library(dplyr)

# Mutate the race variable
working_age <- data_tx %>%
  
  filter(AGE65YR == 1) %>% # filter for ages 18-64

 mutate(race_recode = case_when(
    RACE1 == 1 ~ "White, non-Hispanic",
    RACE1 == 2 ~ "Black, non-Hispanic",
    RACE1 == 3 ~ "American Indian, non-Hispanic",
    RACE1 == 4 ~ "Asian, non-Hispanic",
    RACE1 == 5 ~ "Native Hawaiian or other Pacific Islander, non-Hispanic",
    RACE1 == 7 ~ "Multiracial, non-Hispanic",
    RACE1 == 8 ~ "Hispanic",
    TRUE ~ NA_character_),
    
  obesity = case_when(
    BMI5CAT == 4 ~ "Obese",
    BMI5CAT == 3 ~ "Overweight",
    BMI5CAT == 2 ~ "Normal Weight",
    BMI5CAT == 1 ~ "Underweight",
    TRUE ~ NA_character_
  ),
  
  diabetes = case_when(
    DIABETE4 == 4 ~ "No, pre-diabetes or borderline diabetes",
    DIABETE4 == 3 ~ "No",
    DIABETE4 == 2 ~ "Yes, gestational",
    DIABETE4 == 1 ~ "Yes",
    TRUE ~ NA_character_
  ),
  
  age = case_when(
    AGE65YR == 1 ~ "18 to 64",
    AGE65YR == 2 ~ "65 or older",
    TRUE ~ NA_character_
  
  ))


```

**Step 3: Create survey design. STRATA and WEIGHT statements must be added to take into account**

**the sampling design.**

```{r}
# Create survey design

options(survey.lonely.psu = "adjust")
library(survey)
# Create survey design object
my_design <- svydesign(ids = ~1,
                       strata = ~STSTR,
                       weights = ~LLCPWT,
                       data = working_age)
```

**Step 4: Create a dataframe with diabetes rates by race within the working-age population.**

```{r}

# Use svytable() to create a table of race by diabetes status
diabetes_by_race_table <- svytable(~ race_recode + diabetes, design = my_design)

# Convert the table to a data frame for easier viewing and manipulation, if desired
diabetes_by_race_df <- as.data.frame(diabetes_by_race_table)

# View the table
print(diabetes_by_race_df)


```

**Step 5: Save to Excel.**

```{r}

library(writexl)

write_xlsx(diabetes_by_race_df, path = "Diabetes Rates Working Age by Race 2022.xlsx")

```
